---
title: "HarvardX - Data Science"
author: "Marco Schicker"
date: "2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

## Executive summary

The goal of this project is to predict movie ratings with a low RMSE.
The means to do so are analyzing a data set provided and training a model to make predictions for a validation set of data. 

the edx-data set provided consists of 9000055 movie ratings. Each rating is categorized by the following variables:
* userID - a unique identifier per user
* movieID - a unique identifier per movie
* Title - The movie title, including the release year of the movie
* timestamp - The date and time when the movie was rated by the user
* genres - a list of genres to categorize movies

In general the steps performed to fulfill the task are:
* importing data
* cleaning data
* exploring and visualizing data
* training different models and comparing performance
* validating the most promising model on the validation set

In the end a Model was found with a RMSE of XYZ





## Methods & Analysis
The process to train the algorithm consists of the following basic steps:
* data import
* data cleaning
* data exploration and visualization
* creating training and test set
* 

### data import
The data used to train the model is derived from the publicly available movielens database. According to provided code a data set "edx" and a dataset "validation" is automatically created. 
The "edx"-data set is used to train a model, while the second data set "validation" is used to validate the chosen model. 


### data cleaning
In order to work with the dataset it is analyzed and changed.

* find predictors that are highly correlated and remove if necessary
``` 
corrplot(cor(d), method = "number")
```
* remove predictors with near zero variation
```
List of NZ-predictors and values of STD-deviation
```


### data exploration and visualization
looking at the predictor´s influence on the rating we can see that...
```
grid Y=rating, X=predictors
```


some data about single predictors
* histogram

### insights gained
We can see from the analyzed data that 

### modeling approach
#### identify predictors
#### identify prediction model

## Results
Following the approach mentioned above we can see that we reach the best result by combining the predictor set X (a, b, c, d) with the prediction method Y (ensemble)
```
matrix predictor sets vs prediction method
````


### comparison of model performance


## Conclusion
To sum up the results of this report it was possible to train an algorithm and reach an _RMSE of _.
The best performance was reached by using the following predictors:
* 
* 
*

The best results were reached using an ensemble of the following prediction methods:
* naivebayes
* knn
* random forest
* ...

The limitations of this report lie in...
Next steps to improve performance even more would be to increase training data size, keep on learning or use neural networks to find new predictors in deep learning algorithms.
Use other predictors like actors, country of origin, languages available,...
One could also analyze for trigger words within the title and check if there are correlations with ratings if the title contains words like "Love", "reloaded", "fight" or others.


